PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-21 peace/papers/NAR_web_server_2009/paper.tex
Semi-final changes to the resubmission

\section{\peace\/: Installation and Use}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For
our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
to generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~D.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  \mc{The most important method of quality
  assessment is {\it sensitivity} (the fraction of fragment pairs from
  the same gene that were} \mc{correctly clustered together).  But we also
  look at the {\it Jaccard Index} (which balances sensitivity with the
  number of false positives by looking at the ratio of true positives
  to the sum of true positives, false positives, and true negatives),
  {\it Type 1 error} (the fraction of genes that were divided between
  clusters), and {\it Type 2 error} (the fraction of clusters
  containing two or more genes).  Specificity was not an issue in
  simulated tests, as both \peace\/ and \wcd\/ were completely
  successful in separating unrelated ESTs (save when dealing with
  recently duplicated genes).  In Figure~\ref{SeJiT1T2} we plot
  these four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors too a 17\%
  improvement for 12 processors).}

\mc{We further note, as a point of interest that, we can significantly
  improve the quality of \peace simulation results through the
  increase of the threshold value (see Supplementary Materials,
  Section C) -- achieving a significant improved in PEACE sensitivity
  without an adverse effect on the Jaccard index (i.e. a significant
  increase in the incorrect merging of clusters).  However, the
  improvement does not carry over to the application of real data,
  where we observe a significant drop in the Jaccard Index.  Due to
  this drop we have opted to leave the lower threshold value as the
  default, but users might consider the higher value, thus decreasing
  the risk of incorrectly splitting clusters, if planning on
  using on the subsequent use of a assembly tool that capable of
  breaking up the incorrectly merged clusters during transcript reconstruction.}

@@

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline Bicyclist Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

\mc{We tested the three tools on short-read data using the \metasim\/
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  \metasim\/ are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology.
  Leaving \capthree\/ and \wcd\/ at their default values, but
  employing the {\it adaptive $d^2$} strategy for short read sequences
  (see Section~C in the supplementary materials), we find a that
  \peace\/ performs significantly better than the other two tools on
  454 data using the default \metasim\/ model (with sequence sizes
  ranging in size from 200bp to 370bp, at an average size of 250bp and
  a standard deviation of 17bp).  Over thirty runs, we see an average
  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
  \capthree.  For the Jaccard Index we get essentially the same
  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
  note that it is likely that the \wcd results could be considerably
  improved by the adjustment of a few parameters, but were unable to
  determine how to do that.  (Runtime of \peace\/ was somewhat slower
  than that of \wcd, but given the significant improvement in quality
  of results this is unimportant.)}

@@

>>>>

\section{\peace: Installation and Use}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For
our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
to generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~D.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  \mc{The most important method of quality
  assessment is {\it sensitivity} (the fraction of fragment pairs from
  the same gene that were} \mc{correctly clustered together).  But we
  also look at the {\it Jaccard Index} (which balances sensitivity
  with the number of false positives by looking at the ratio of true
  positives to the sum of true positives, false positives, and true
  negatives), {\it Type 1 error} (the fraction of genes that were
  divided between clusters), and {\it Type 2 error} (the fraction of
  clusters containing two or more genes).  Specificity was not an
  issue in simulated tests, as both \peace\/ and \wcd\/ were
  completely successful in separating unrelated ESTs (with the
  exception of recently duplicated genes, as discussed in the
  Supplementary Materials).  In Figure~\ref{SeJiT1T2} we plot these
  four tests as a function of error rate, and observe the almost
  identical results between \peace\/ and \wcd\/.  In
  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
  \wcd\/, again observing almost identical results when run
  sequentially -- but significantly faster runtime for PEACE on
  multiple processor when holding the EST/processor ratio constant
  (ranging from a 65\% improvement for two processors too a 17\%
  improvement for 12 processors).}

\mc{In the course of our simulations we also invetigated the memory
  footprint -- of particular concern for large data sets.  We find
  that, as with \wcd, the required memory for an execution is linear
  in the size of the dataset.}

  \mc{We note, as a point of interest, that we can significantly
    improve the quality of \peace\/ simulation results through the
    increase of the threshold value (see Supplementary Materials,
    Section C) -- achieving a significant improved in PEACE
    sensitivity without an adverse effect on the Jaccard index (i.e. a
    significant increase in the incorrect merging of clusters).
    However, the improvement does not carry over to the application of
    real data (Supplementary Materials, Section E.2), where we observe
    a significant increase in the incorrect merging of clusters.
    While this might be acceptable to a user planning to employ an
    assembl capbable of breaking up the clusters, we have choosed to
    base our analysis (and set out default values) at the lower threshold.}

@@

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

\mc{We tested the three tools on short-read data using the \metasim\/
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  \metasim\/ are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology
  (producing reads ranging in size from 200bp to 370bp with a mean of
  250bp and a standard deviation of 17bp), and the Illumina short-read
  technology (producing reads of exactly 62bp).  Experiements were run
  using two sets of \wcd\/ parameters (the default paramters, and
  setting window size to 75 and threshold to 75 -- thus matching the
  \peace\/ parameters for this size squence), and using default values
  of \capthree.  Only with \peace were we able to get reasonable
  results, seeing a sensitivity of 0.871 for the 454 reads and
  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
  0.372 respectivally, and other comperable figures.  (See Table~S3 in
  the Supplementary Materials.)  In short, \peace performs quite well
  for 454 reads and provides some useful information for Illumina
  reads, while \wcd\. and \capthree are unable to produce working
  results.}

@@

Peace can be downloaded from
\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
committed to keep maintaining and improving the tool in the future.
Meanwhile, we are developing our own assembly tool that can seamlessly
integrate with PEACE.  The underlying modular design of PEACE offers
users many possibilities to expand and incorporate the MST algorithm
for other bioinformatics application.
